---
title: "Correcting for spatially biased sampling in compositional data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correcting for spatially biased sampling in compositional data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Fisheries catch composition data are often collected from non-representative
sampling programmes — observers may cover only a subset of the fleet, certain
areas or seasons may be over- or under-sampled, and spatial coverage can shift
over time. Naively treating these samples as representative of the whole fishery
can produce biased estimates of catch-at-age or catch-at-length compositions.

This vignette demonstrates how `ComPoM` can correct for such biases by:
fitting a Poisson-multinomial model with spatially varying coefficients
(via `sdmTMB`), then scaling the model predictions to known total catches
across the full fished area.

The simulation sets up a known truth — spatially varying selectivity, temporal
trends, and seasonal effects — then samples it with strong spatial and seasonal
bias. We compare a spatial model that accounts for the full domain against a
non-spatial model that cannot, showing that the spatial model recovers the true
compositions more accurately.

```{r setup}
library(tidyverse)
library(ComPoM)
library(sdmTMB)

source(here::here("R/helpers.R"))

# Helper functions
clo <- function(x) x / sum(x)

# Simulation parameters
n_bins <- 10
n_samps <- 50
n_locations <- 250
n_seasons <- 4
n_per_samps <- 100
n_years <- 10

M <- 0.2

surv <- exp(-(1:n_bins) * M)
popfreq <- clo(surv)

# Selectivity parameters
L50_base <- 6.5
L95 <- 0.75
```

## Simulate Spatial Domain

Create a spatial grid with coordinates and simulate spatially-varying selectivity.

```{r simulate-spatial-domain, fig.width=7, fig.height=7}
set.seed(123)

# Create spatial locations (e.g., fishing locations)
locations <- data.frame(
  loc_id = 1:n_locations,
  x = runif(n_locations, 0, 100),
  y = runif(n_locations, 0, 100)
)

# Simulate spatial random field for selectivity variation
# Using a simple distance-based correlation structure
spatial_field <- function(x, y, range = 30, sigma = 0.3) {
  n <- length(x)
  dist_mat <- as.matrix(dist(cbind(x, y)))
  cov_mat <- sigma^2 * exp(-dist_mat / range)
  MASS::mvrnorm(1, rep(0, n), cov_mat)
}

# Generate spatial effect on L50
locations$spatial_effect <- spatial_field(locations$x, locations$y, range = 10, sigma = 0.5)

# Visualize spatial field
ggplot(locations, aes(x = x, y = y, color = spatial_effect)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial effect on selectivity (L50)")
```

## Simulate Temporal and Seasonal Effects

```{r simulate-temporal-effects}
# Year effects (random walk)
year_effects <- cumsum(rnorm(n_years, 0, 0.1))

# Season effects
season_effects <- rnorm(n_seasons, 0, 0.2)
season_effects <- season_effects - mean(season_effects)

temporal_df <- expand_grid(
  Year = factor(1:n_years),
  Season = factor(1:n_seasons)
) |>
  mutate(
    year_effect = year_effects[as.integer(Year)],
    season_effect = season_effects[as.integer(Season)]
  )
```

## Generate True Catch Compositions

```{r true-catch-compositions}
# Full factorial of all strata with spatial coordinates
true_catch_by_bin <- expand_grid(
  loc_id = 1:n_locations,
  Year = factor(1:n_years),
  Season = factor(1:n_seasons),
  bin = 1:n_bins
) |>
  left_join(locations, by = "loc_id") |>
  left_join(temporal_df, by = c("Year", "Season")) |>
  mutate(
    L50 = exp(log(L50_base) + spatial_effect + year_effect + season_effect),
    catch = rgamma(n(), 100, 0.01),
    tcatch = Selectivity(bin, L50, L95) * popfreq[bin] * catch
  )

# Summarize true catch
true_catch_summary <- true_catch_by_bin |>
  group_by(Year, bin) |>
  summarise(tcatch = sum(tcatch), .groups = "drop")
```

## Simulate Non-Representative Sampling

```{r simulate-sampling, fig.width=7, fig.height=7}
# Sample locations with strong spatial bias (prefer high x AND high y corner)
sample_locations <- function(n_samps, locations) {
  # Strong bias towards the northeast corner of the domain
  probs <- exp(locations$x / 25 + locations$y / 25)
  probs <- probs / sum(probs)
  sample(locations$loc_id, n_samps, replace = TRUE, prob = probs)
}

# Season sampling probabilities that shift across years
# Early years oversample season 1, late years oversample season 4
season_probs_by_year <- function(year, n_years, n_seasons) {
  # Linear shift: weight moves from season 1 to season 4 over time
  alpha <- (year - 1) / (n_years - 1)  # 0 for year 1, 1 for last year
  weights <- numeric(n_seasons)
  weights[1] <- 4 * (1 - alpha) + 0.5 * alpha

  weights[n_seasons] <- 0.5 * (1 - alpha) + 4 * alpha
  # Middle seasons stay low
  if (n_seasons > 2) {
    weights[2:(n_seasons - 1)] <- 0.5
  }
  weights / sum(weights)
}

# Generate samples - distribute across years
set.seed(456)
samples <- tibble(
  samp_id = 1:(n_samps * n_years),
  Year = factor(rep(1:n_years, each = n_samps))
) |>
  mutate(
    loc_id = sample_locations(n(), locations),
    Season = factor(sapply(as.integer(Year), function(yr) {
      sample(1:n_seasons, 1, prob = season_probs_by_year(yr, n_years, n_seasons))
    }))
  ) |>
  left_join(locations, by = "loc_id") |>
  left_join(temporal_df, by = c("Year", "Season")) |>
  mutate(
    L50 = exp(log(L50_base) + spatial_effect + year_effect + season_effect)
  )

# Generate composition samples from each sampling event
long_form_sims <- samples |>
  group_by(samp_id) |>
  nest() |>
  mutate(
    comp_samps = lapply(data, function(df) {
      which(rmultinom(n_per_samps, 1,
                      Selectivity(1:n_bins, df$L50, L95) * popfreq) == 1,
            arr.ind = TRUE)[, 1]
    })
  ) |>
  unnest(c(data, comp_samps)) |>
  mutate(count = 1) |>
  rename(bin = comp_samps)

# Visualize sampling locations
ggplot() +
  geom_point(data = locations, aes(x = x, y = y), color = "grey70", size = 2) +
  geom_point(data = long_form_sims |> distinct(loc_id, x, y),
             aes(x = x, y = y), color = "red", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Sampling locations (red) vs all locations (grey)")
```

## Prepare Data and Create Mesh

```{r prepare-data-mesh, fig.width=5, fig.height=4}
# Prepare input data for ComPoM
input_data <- data_prep(
  long_form_sims,
  vars_for_grouping = c('loc_id', 'Year', 'Season', 'x', 'y'),
  bin_lab = 'bin'
) |>
  mutate(
    x = as.numeric(as.character(x)),
    y = as.numeric(as.character(y))
  )

# Create mesh for sdmTMB — use ALL fishing locations for the domain,
# not just the biased sample, so the mesh covers the full fished area
all_locs_sf <- sf::st_as_sf(locations, coords = c("x", "y"))
bnd <- fmesher::fm_nonconvex_hull(all_locs_sf, convex = -0.1, format = 'sf')
msh <- fmesher::fm_mesh_2d(locations[, c("x", "y")], cutoff = 3, boundary = bnd)
mesh <- make_mesh(input_data, xy_cols = c("x", "y"), mesh = msh)

bnd_vect <- terra::vect(bnd)
pg <- terra::rast(bnd_vect)
pg$lyr.1 <- 1

plot(mesh)
```

## Fit Spatiotemporal Model with sdmTMB

```{r fit-spatiotemporal-model}
# Fit model with bin-specific spatially varying coefficients
mods_spatial <- fit_model(
  re_form = 'Season',
  ffx_form = 'Year',
  data = input_data,
  backend = 'TMB',
  mesh = mesh,
  time = "Year",
  coords = c("x", "y"),
  iter = 1000,
  warmup = 500,
  thin = 2
)

summary(mods_spatial$model)
```

## Fit Non-Spatial Model for Comparison

```{r fit-nonspatial-model}
mods_nonspatial <- fit_model(
  re_form = 'Season',
  ffx_form = 'Year',
  data = input_data,
  backend = 'TMB',
  iter = 1000,
  warmup = 500,
  thin = 2
)

summary(mods_nonspatial$model)
```

## Compare Model Fits

```{r compare-aic}
AIC(mods_spatial$model)
AIC(mods_nonspatial$model)
```

## Effects Plots

```{r effects-plots, fig.dim=c(12,7)}
Fx_plot(mods_nonspatial, grp = 'Year', grid = NULL) +
  ggtitle("Year effects (Non-spatial model)")

Fx_plot(mods_spatial, grp = 'Year', grid = NULL) +
  ggtitle("Year effects (Spatial model)")

Fx_plot(mods_nonspatial, grp = 'Season', grid = NULL) +
  ggtitle("Season effects (Non-spatial model)")

Fx_plot(mods_spatial, grp = 'Season', grid = NULL) +
  ggtitle("Season effects (Spatial model)")
```

## Scale Compositions to Catch

```{r scale-compositions, fig.dim=c(12,7)}
# Prepare scaling data for spatial model
IDs <- terra::extract(pg, true_catch_by_bin[, c('x', 'y')], ID = TRUE, xy = TRUE)
true_catch_by_bin[, colnames(IDs)] <- IDs

scale_df_spatial <- true_catch_by_bin |>
  group_by(Season, Year, x, y) |>
  summarise(catch = sum(tcatch), .groups = "drop")

# Prepare scaling data for non-spatial model
scale_df_nonspatial <- true_catch_by_bin |>
  semi_join(
    input_data |> ungroup() |> distinct(Year, Season),
    by = c("Year", "Season")
  ) |>
  group_by(Year, Season) |>
  summarise(catch = sum(tcatch), .groups = "drop")

# Scale compositions
scaled_comp_spatial <- scale_comps(
  scale_df_spatial,
  predvar = 'catch',
  fit = mods_spatial,
  grps = c('Season', 'Year')
)

scaled_comp_nonspatial <- scale_comps(
  scale_df_nonspatial,
  predvar = 'catch',
  fit = mods_nonspatial,
  grps = c('Year', 'Season')
)
```

## Compare Scaled Compositions to Truth

```{r compare-compositions, fig.dim=c(12,7)}
# True compositions by year
true_comp_year <- true_catch_by_bin |>
  group_by(Year, bin) |>
  summarise(tcatch = sum(tcatch), .groups = "drop")

# Spatial model (aggregate across seasons)
scaled_comp_spatial_yr <- scaled_comp_spatial |>
  group_by(Year, bin, .draw) |>
  summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

scaled_comp_plot(
  scaled_comp = scaled_comp_spatial_yr,
  grps = 'Year',
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  cvar = 'tcatch'
) + ggtitle("Spatial model vs true compositions (by year)")

# Non-spatial model (aggregate across seasons)
scaled_comp_nonspatial_yr <- scaled_comp_nonspatial |>
  group_by(Year, bin, .draw) |>
  summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

scaled_comp_plot(
  scaled_comp = scaled_comp_nonspatial_yr,
  grps = 'Year',
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  cvar = 'tcatch'
) + ggtitle("Non-spatial model vs true compositions (by year)")
```

## Evaluate Model Performance

```{r evaluate-performance, fig.dim=c(12,7)}
# Raw (biased) samples
raw_comp <- input_data |>
  group_by(Year, bin) |>
  summarise(tcatch = sum(tot_by_bin), .groups = "drop")

# Spatial model comparison
scaled_comp_plot(
  scaled_comp = scaled_comp_spatial_yr,
  grps = 'Year',
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = 'tcatch'
) + ggtitle("Blue = True, Orange = Raw samples, Black = Spatial model")

scaled_comp_comp(
  scaled_comp = scaled_comp_spatial_yr,
  grps = 'Year',
  scaled = TRUE,
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = 'tcatch'
) + ggtitle("Spatial model vs raw samples (by year)")

# Non-spatial model comparison
scaled_comp_plot(
  scaled_comp = scaled_comp_nonspatial_yr,
  grps = 'Year',
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = 'tcatch'
) + ggtitle("Blue = True, Orange = Raw samples, Black = Non-spatial model")
```

## Model Diagnostics

```{r model-diagnostics}
tidy(mods_spatial$model, "ran_pars")
```

The spatial model can better account for the non-representative sampling by modeling the underlying spatial structure in the data.

## Oceanographic Influence on Compositions

Environmental covariates such as sea surface temperature (SST) can drive
spatial and temporal shifts in fish size compositions independently of the
year and season effects of primary interest. A recurring oceanographic cycle
such as El Niño — warm SST every ~5 years — shifts fish size availability:
warm years have larger fish more available; cold years have smaller fish.

Including SST as a smooth spline (`spl = "by_factor"`) separates the
oceanographic signal from the year fixed effects. `scale_comps()` can then
either retain the SST effect (compositions as observed under the ENSO cycle)
or remove it by evaluating the spline at mean SST (`sst_c = 0`), giving
year compositions standardised to average oceanographic conditions.

### Simulate ENSO-Like SST Cycle

We simulate a 5-year sinusoidal SST cycle (peaks at years 2, 7; troughs at
years 5, 10) plus independent spatial noise. The spatial noise is zero-centred
within each year so the within-year mean SST equals the cycle value — SST
carries both a temporal trend and within-year spatial variation.

```{r sst-simulate}
set.seed(42)

# 5-year sinusoidal ENSO cycle: warm years 2, 3, 7, 8 — cold years 4, 5, 9, 10
sst_cycle <- 1.5 * sin(2 * pi * (1:n_years - 1) / 5)
cat("SST cycle by year:", round(sst_cycle, 2), "\n")

# Spatial SST: temporal cycle + zero-centred spatial noise per year
sst_loc_year_enso <- purrr::map_dfr(1:n_years, function(yr) {
  spatial_noise <- spatial_field(locations$x, locations$y, range = 30, sigma = 0.4) +
                   rnorm(nrow(locations), 0, 0.1)
  locations |>
    dplyr::mutate(
      Year     = factor(yr),
      sst_spat = spatial_noise - mean(spatial_noise),
      sst_c    = sst_cycle[yr] + sst_spat
    )
})

# Verify within-year SST mean tracks the cycle
sst_loc_year_enso |>
  dplyr::group_by(Year) |>
  dplyr::summarise(mean_sst = round(mean(sst_c), 2),
                   sd_sst   = round(sd(sst_c),   2)) |>
  dplyr::mutate(cycle = round(sst_cycle, 2))
```

### Simulate SST-Shifted Compositions

Warm SST shifts L50 upward — larger fish become more available.

```{r sst-resample}
set.seed(42)

L50_sst_base <- 5
L95_sst      <- 8

input_sst_enso <- true_catch_by_bin |>
  dplyr::left_join(
    sst_loc_year_enso |> dplyr::select(loc_id, Year, sst_c),
    by = c("loc_id", "Year")
  ) |>
  dplyr::mutate(
    L50_use = L50_sst_base + 1.5 * sst_c,
    sel_use = Selectivity(bin, L50_use, L95_sst),
    avail   = popfreq[bin] * sel_use
  ) |>
  dplyr::group_by(loc_id, Year, Season) |>
  dplyr::mutate(
    prob   = avail / sum(avail),
    counts = as.integer(rmultinom(1, n_per_samps, prob))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(loc_id, Year, Season, bin, counts, sst_c)

# Verify: proportion of large fish tracks the ENSO cycle
input_sst_enso |>
  dplyr::group_by(Year) |>
  dplyr::summarise(
    prop_large = round(sum(counts[bin >= 6]) / sum(counts), 3),
    mean_sst   = round(mean(sst_c), 2)
  )
```

### Aggregate to SST Quintile Strata

SST quintiles are computed within each year, preserving both the temporal
cycle and within-year spatial variation independently of the year dummies.

```{r sst-aggregate}
input_sst_binned_enso <- input_sst_enso |>
  dplyr::group_by(Year) |>
  dplyr::mutate(
    sst_q   = dplyr::ntile(sst_c, 5),
    sst_mid = tapply(sst_c, sst_q, mean)[as.character(sst_q)]
  ) |>
  dplyr::ungroup() |>
  dplyr::group_by(Year, Season, sst_q, sst_mid, bin) |>
  dplyr::summarise(counts = sum(counts), .groups = "drop")

prep_sst_enso <- data_prep(
  comp              = input_sst_binned_enso |> dplyr::mutate(sst_c = sst_mid),
  vars_for_grouping = c("Year", "Season", "sst_q", "sst_c"),
  bin_lab           = "bin",
  count_var         = "counts"
)

cat("Unique SST values:", length(unique(prep_sst_enso$sst_c)), "\n")
```

### Fit Model with SST Smooth

`spl = "by_factor"` fits `s(sst_c, by=bin, k=4)` — a separate 1D smooth of
SST for each bin level, capturing bin-specific responses to SST (large bins
increase more in warm years). This is full-rank alongside `factor(Year):bin`
because SST varies across quintile strata within each year independently of
the year dummies. `Season` enters as nuisance RE via `re_form`.

```{r sst-fit}
mods_sst_enso <- suppressWarnings(fit_model(
  ffx_form = "Year",
  re_form  = "Season",
  cvars    = "sst_c",
  spl      = "by_factor",
  knots    = "4",
  data     = prep_sst_enso,
  backend  = "TMB",
  iter     = 1000,
  warmup   = 500,
  thin     = 2
))

mods_sst_enso$model <- sdmTMB::run_extra_optimization(
  mods_sst_enso$model, newton_loops = 2
)
sanity(mods_sst_enso$model)
```

### Year Effects After Accounting for SST

```{r sst-fx-plot, fig.dim=c(12, 5)}
Fx_plot(mods_sst_enso, grp = "Year") +
  ggtitle("Year-effect multipliers after accounting for ENSO SST spline")
```

### Scale Compositions — Retaining vs Removing ENSO Signal

`remove_nuisance = FALSE` evaluates the SST spline at the observed quintile
midpoint per stratum. Warm years (2, 7) produce larger-fish-dominated
compositions reflecting the true ENSO-driven availability.

`remove_nuisance = TRUE` with `sst_c = 0` evaluates the spline at mean SST
and drops the Season RE, giving year compositions standardised to average
oceanographic conditions and removing the ENSO signal.

```{r sst-scale}
true_comp_sst_enso <- input_sst_binned_enso |>
  dplyr::group_by(Year, bin) |>
  dplyr::summarise(tcatch = sum(counts), .groups = "drop")

# Catch apportioned equally across 5 SST quintiles within each Year x Season
scale_df_sst_enso <- input_sst_binned_enso |>
  dplyr::distinct(Year, Season, sst_q, sst_c = sst_mid) |>
  dplyr::left_join(
    scale_df_nonspatial |> dplyr::select(Year, Season, catch),
    by = c("Year", "Season")
  ) |>
  dplyr::mutate(catch = catch / 5)

# ENSO signal retained
scaled_enso_with <- suppressWarnings(scale_comps(
  scale_df        = scale_df_sst_enso,
  predvar         = "catch",
  fit             = mods_sst_enso,
  grps            = c("Year", "Season", "sst_q"),
  remove_nuisance = FALSE
))

# ENSO signal removed: SST set to 0 (mean), Season RE marginalised out
scaled_enso_without <- suppressWarnings(scale_comps(
  scale_df        = scale_df_sst_enso |> dplyr::mutate(sst_c = 0),
  predvar         = "catch",
  fit             = mods_sst_enso,
  grps            = c("Year", "Season", "sst_q"),
  remove_nuisance = TRUE
))
```

### Compare Annual Compositions

```{r sst-compare, fig.dim=c(12, 10)}
enso_with_yr <- scaled_enso_with |>
  dplyr::group_by(Year, bin, .draw) |>
  dplyr::summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

enso_without_yr <- scaled_enso_without |>
  dplyr::group_by(Year, bin, .draw) |>
  dplyr::summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

p_enso_with <- scaled_comp_plot(
  scaled_comp = enso_with_yr, grps = "Year", scaled = TRUE,
  comp_are    = true_comp_sst_enso, cvar = "tcatch"
) + ggtitle("SST effect retained (observed ENSO cycle)")

p_enso_without <- scaled_comp_plot(
  scaled_comp = enso_without_yr, grps = "Year", scaled = TRUE,
  comp_are    = true_comp_sst_enso, cvar = "tcatch"
) + ggtitle("SST effect removed (standardised to mean SST = 0)")

cowplot::plot_grid(p_enso_with, p_enso_without, nrow = 2)
```

With SST retained (top), warm ENSO years (2, 7) show relatively more large
fish and cold years (4, 5, 9, 10) show small-fish-dominated compositions,
matching the simulated truth (blue dots). With SST removed (bottom),
predictions converge toward the mean-SST composition — the ENSO-driven
inter-annual variability is standardised out, isolating the residual year
fixed effects from the oceanographic forcing.
