---
title: "Correcting for vessel-level sampling bias in compositional data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correcting for vessel-level sampling bias in compositional data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Observer programmes rarely sample all vessels equally. Some vessels are sampled
in every season, others only sporadically, and the mix of observed vessels
changes from year to year. Because vessels differ in the size-composition of
their catch — through gear configuration, skipper behaviour, or preferred
fishing grounds — the raw observer sample can be a poor representation of the
fleet-wide catch composition.

This vignette demonstrates how `ComPoM` can correct for vessel-level sampling
bias using a Poisson-multinomial model fitted with either a Bayesian backend
(`brms`) or a frequentist random-effects backend (`sdmTMB` without a mesh).
Neither model uses spatial information; instead, both treat vessel identity and
season as crossed random effects interacting with the composition bins.

The simulation generates a fleet of vessels with vessel-specific selectivity
offsets and year/season effects, then samples that fleet unevenly — a small
subset of vessels is heavily over-represented, and their seasonal coverage
shifts across years. We compare three approaches for estimating fleet-wide
catch compositions:

1. **Raw scaled samples** — naively treating observer data as representative
2. **sdmTMB (non-spatial)** — maximum-likelihood Poisson-multinomial with
   random effects
3. **brms (Bayesian)** — fully Bayesian Poisson-multinomial with the same
   random-effect structure

## Setup

```{r setup}
library(tidyverse)
library(ComPoM)
library(sdmTMB)

source(here::here("R/helpers.R"))

clo <- function(x) x / sum(x)

# Simulation parameters
n_bins <- 10
n_vessels <- 40
n_seasons <- 4
n_years <- 8
n_samps_per_year <- 60
n_per_samp <- 100

# Population and selectivity
M <- 0.2
surv <- exp(-(1:n_bins) * M)
popfreq <- clo(surv)

L50_base <- 6.5
L95 <- 0.75
```

## Simulate Fleet

Each vessel has a persistent selectivity offset drawn from a normal
distribution — some vessels catch larger fish, others smaller.

```{r simulate-fleet, fig.width=6, fig.height=4}
set.seed(42)

vessels <- tibble(
  vessel_id = factor(1:n_vessels),
  vessel_effect = rnorm(n_vessels, 0, 0.3)
)

ggplot(vessels, aes(x = vessel_id, y = vessel_effect)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "Vessel-specific selectivity offsets",
    x = "Vessel", y = "Effect on log(L50)"
  )
```

## Simulate Temporal and Seasonal Effects

```{r simulate-temporal}
year_effects <- cumsum(rnorm(n_years, 0, 0.1))
year_effects <- year_effects - mean(year_effects)

season_effects <- rnorm(n_seasons, 0, 0.2)
season_effects <- season_effects - mean(season_effects)

temporal_df <- expand_grid(
  Year = factor(1:n_years),
  Season = factor(1:n_seasons)
) |>
  mutate(
    year_effect = year_effects[as.integer(Year)],
    season_effect = season_effects[as.integer(Season)]
  )
```

## Generate True Catch Compositions

The true fleet-wide composition integrates over all vessels, years, and
seasons.

```{r true-catch}
true_catch_by_bin <- expand_grid(
  vessel_id = factor(1:n_vessels),
  Year = factor(1:n_years),
  Season = factor(1:n_seasons),
  bin = 1:n_bins
) |>
  left_join(vessels, by = "vessel_id") |>
  left_join(temporal_df, by = c("Year", "Season")) |>
  mutate(
    L50 = exp(log(L50_base) + vessel_effect + year_effect + season_effect),
    catch = rgamma(n(), 100, 0.01),
    tcatch = Selectivity(bin, L50, L95) * popfreq[bin] * catch
  )

true_comp_year <- true_catch_by_bin |>
  group_by(Year, bin) |>
  summarise(tcatch = sum(tcatch), .groups = "drop")
```

## Simulate Non-Representative Sampling

Sampling is biased in two ways:

- **Vessel bias**: a handful of vessels (the top 8) are sampled much more
  frequently than the rest.
- **Seasonal shift**: early years oversample season 1; late years oversample
  season 4 — mimicking a programme that changes its coverage over time.

```{r simulate-sampling, fig.width=7, fig.height=5}
# Vessel sampling weights — top 8 vessels get 5x the weight
vessel_weights <- rep(1, n_vessels)
vessel_weights[1:8] <- 5
vessel_weights <- vessel_weights / sum(vessel_weights)

# Seasonal weights that shift over time
season_probs_by_year <- function(year, n_years, n_seasons) {
  alpha <- (year - 1) / (n_years - 1)
  weights <- rep(0.5, n_seasons)
  weights[1] <- 4 * (1 - alpha) + 0.5 * alpha
  weights[n_seasons] <- 0.5 * (1 - alpha) + 4 * alpha
  weights / sum(weights)
}


samples <- tibble(
  samp_id = 1:(n_samps_per_year * n_years),
  Year = factor(rep(1:n_years, each = n_samps_per_year))
) |>
  mutate(
    vessel_id = factor(sample(1:n_vessels, n(), replace = TRUE, prob = vessel_weights)),
    Season = factor(sapply(as.integer(Year), function(yr) {
      sample(1:n_seasons, 1, prob = season_probs_by_year(yr, n_years, n_seasons))
    }))
  ) |>
  left_join(vessels, by = "vessel_id") |>
  left_join(temporal_df, by = c("Year", "Season")) |>
  mutate(
    L50 = exp(log(L50_base) + vessel_effect + year_effect + season_effect)
  )

# Generate composition samples
long_form_sims <- samples |>
  group_by(samp_id) |>
  nest() |>
  mutate(
    comp_samps = lapply(data, function(df) {
      which(rmultinom(n_per_samp, 1,
                      Selectivity(1:n_bins, df$L50, L95) * popfreq) == 1,
            arr.ind = TRUE)[, 1]
    })
  ) |>
  unnest(c(data, comp_samps)) |>
  mutate(count = 1) |>
  rename(bin = comp_samps)

# Visualise sampling bias
sampling_summary <- long_form_sims |>
  distinct(samp_id, vessel_id, Year, Season) |>
  count(vessel_id, name = "n_trips")

ggplot(sampling_summary, aes(x = vessel_id, y = n_trips)) +
  geom_col() +
  theme_minimal() +
  labs(
    title = "Number of sampled trips per vessel",
    x = "Vessel", y = "Trips sampled"
  )
```

```{r sampling-by-season, fig.width=7, fig.height=4}
season_summary <- long_form_sims |>
  distinct(samp_id, Year, Season) |>
  count(Year, Season)

ggplot(season_summary, aes(x = Season, y = n, fill = Season)) +
  geom_col() +
  facet_wrap(~Year, nrow = 1) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Samples per season across years",
    x = "Season", y = "Samples"
  )
```

## Prepare Data

```{r prepare-data}
input_data <- data_prep(
  long_form_sims,
  vars_for_grouping = c("vessel_id", "Year", "Season"),
  bin_lab = "bin"
)
```

## Fit Non-Spatial sdmTMB Model

```{r fit-tmb}
mod_tmb <- fit_model(
  re_form = "Season + vessel_id",
  ffx_form = "Year",
  data = input_data,
  backend = "TMB",
  iter = 1000,
  warmup = 500,
  thin = 2
)
```

## Fit Bayesian Model with brms

```{r fit-brms, message=FALSE}
mod_brms <- fit_model(
  re_form = "Season + vessel_id",
  ffx_form = "Year",
  data = input_data,
  backend = "brms",
  dist=poisson(),
  brms_backend = "cmdstanr",
  chains = 4,
  cores = 4,
  threads = 1,
  iter = 1000,
  warmup = 500,
  thin = 2
)
```

## Effects Plots

```{r effects-year, fig.dim=c(10, 6)}
Fx_plot(mod_tmb, grp = "Year", grid = NULL) +
  ggtitle("Year effects (sdmTMB)")

Fx_plot(mod_brms, grp = "Year", grid = NULL) +
  ggtitle("Year effects (brms)")
```

```{r effects-season, fig.dim=c(10, 5)}
Fx_plot(mod_tmb, grp = "Season", grid = NULL) +
  ggtitle("Season effects (sdmTMB)")

Fx_plot(mod_brms, grp = "Season", grid = NULL) +
  ggtitle("Season effects (brms)")
```

## Scale Compositions to Catch

We scale predictions from each model to the true total catch by year and
season (summed across all vessels), then aggregate to annual compositions.

```{r scale-compositions}
# Scaling data: total catch by year and season across all vessels
scale_df <- true_catch_by_bin |>
  group_by(Year, Season, vessel_id) |>
  summarise(catch = sum(tcatch), .groups = "drop")

scaled_tmb <- scale_comps(
  scale_df,
  predvar = "catch",
  fit = mod_tmb,
  grps = c("Year", "Season", "vessel_id")
)

scaled_brms <- scale_comps(
  scale_df,
  predvar = "catch",
  fit = mod_brms,
  grps = c("Year", "Season", "vessel_id")
)

# Aggregate to annual
scaled_tmb_yr <- scaled_tmb |>
  group_by(Year, bin, .draw) |>
  summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

scaled_brms_yr <- scaled_brms |>
  group_by(Year, bin, .draw) |>
  summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")
```

## Compare Scaled Compositions to Truth

```{r compare-truth-tmb, fig.dim=c(10, 7)}
scaled_comp_plot(
  scaled_comp = scaled_tmb_yr,
  grps = "Year",
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  cvar = "tcatch"
) + ggtitle("sdmTMB vs true compositions (by year)")
```

```{r compare-truth-brms, fig.dim=c(10, 7)}
scaled_comp_plot(
  scaled_comp = scaled_brms_yr,
  grps = "Year",
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  cvar = "tcatch"
) + ggtitle("brms vs true compositions (by year)")
```

## Evaluate Against Raw Samples

```{r raw-comp}
raw_comp <- input_data |>
  group_by(Year, bin) |>
  summarise(tcatch = sum(tot_by_bin), .groups = "drop")
```

```{r compare-raw-tmb, fig.dim=c(10, 7)}
scaled_comp_plot(
  scaled_comp = scaled_tmb_yr,
  grps = "Year",
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = "tcatch"
) + ggtitle("Blue = True, Orange = Raw, Black = sdmTMB")
```

```{r compare-raw-brms, fig.dim=c(10, 7)}
scaled_comp_plot(
  scaled_comp = scaled_brms_yr,
  grps = "Year",
  scaled = TRUE,
  scales = "free_y",
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = "tcatch"
) + ggtitle("Blue = True, Orange = Raw, Black = brms")
```

## Ridgeline Plots of Scaled Compositions

Ridgeline plots display how the modelled size composition shifts across years,
making temporal trends easier to read than a grid of point-range panels.
`scaled_ridge_plot()` takes the output of `scale_comps()` directly; `plot_y`
sets the variable stacked on the y-axis and `plot_facet` optionally splits the
display across a second variable (here, not needed as we are already aggregated
to annual).

```{r ridge-tmb, fig.dim=c(8, 7)}
library(ggridges)

scaled_ridge_plot(
  scaled_comp = scaled_tmb_yr,
  grps        = "Year",
  plot_y      = "Year"
) + ggtitle("sdmTMB — annual size compositions (ridgeline)")
```

```{r ridge-brms, fig.dim=c(8, 7)}
scaled_ridge_plot(
  scaled_comp = scaled_brms_yr,
  grps        = "Year",
  plot_y      = "Year"
) + ggtitle("brms — annual size compositions (ridgeline)")
```

If you have a second grouping variable (e.g. season within year) you can pass
it to `plot_facet` to produce a faceted ridgeline display:

```{r ridge-tmb-season, fig.dim=c(10, 8)}
# Re-aggregate retaining Season
scaled_tmb_yr_season <- scaled_tmb |>
  dplyr::group_by(Year, Season, bin, .draw) |>
  dplyr::summarise(tot_by_bin = sum(tot_by_bin), .groups = "drop")

scaled_ridge_plot(
  scaled_comp = scaled_tmb_yr_season,
  grps        = c("Year", "Season"),
  plot_y      = "Year",
  plot_facet  = "Season"
) + ggtitle("sdmTMB — annual compositions faceted by season")
```

## Relative Error Comparison

```{r relative-error-tmb, fig.dim=c(8, 5)}
scaled_comp_comp(
  scaled_comp = scaled_tmb_yr,
  grps = "Year",
  scaled = TRUE,
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = "tcatch"
) + ggtitle("sdmTMB model vs raw samples — relative error")
```

```{r relative-error-brms, fig.dim=c(8, 5)}
scaled_comp_comp(
  scaled_comp = scaled_brms_yr,
  grps = "Year",
  scaled = TRUE,
  comp_are = true_comp_year,
  comp_are2 = raw_comp,
  cvar = "tcatch"
) + ggtitle("brms model vs raw samples — relative error")
```

Both the Bayesian and frequentist models should recover the true fleet-wide
compositions more accurately than the raw biased samples, particularly in years
where the sampled vessel mix or seasonal coverage departs most from the fleet
average.
